Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-16 treestats_paper.tex
Rejigged the abstract.

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Summaries of genetic variation therefore also describe the underlying genealogies.
Here we exploit this duality to create a method that efficiently computes single-site statistics
from population genetic data using the succinct tree sequence encoding
of genealogies and genome sequence.
We use the framework to provide efficient implementations of
many currently-defined statistics, and to suggest several new statistics.
The general approach aggregates ``sample weights'' up each marginal tree,
which are then summarised using a ``summary function'';
different statistics result from different choices of weight or function.
The results can be summarised in three different ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence;
by \emph{branch}, which gives the expected value of the site statistic
with a large number of mutations under the infinite-sites model;
or by \emph{node}, which summarises the contribution of each node in the tree to the statistics.
This formulation also makes it apparent what statistic of the underlying genealogical trees
each sequence-based statistic is estimating.  By using simulations to evaluate performance, we
show that calculating statistics from tree sequences is several orders of magnitude more efficient than
methods traversing the entire data matrix in terms of both run time and memory requirements.
We also use tree sequences inferred from the 1000 Genomes Project dataset,
to evaluate how well this duality between site and branch statistics holds
and to describe the temporal origin of some admixture statistics.

>>>>

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We exploit this duality here to create a method that efficiently computes single-site statistics
from population genetic data using the succinct tree sequence encoding
of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two (dual) ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
with a large number of mutations under the infinite-sites model.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
bespoke matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may hold interesting biological
signals.

\jk{Not sure if the last bit is expressed the right way}
